## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
x <- 10
if (x > 5) {
  print("x is greater than 5")
}

```

The logical statements should boil down to a boolean; for example:

```{r}
z <- 15
result <- (x < y) & (z > y)  
```

We can also add else if statements in the middle if we need more options.

```{r}
grade <- 85

if (grade >= 90) {
  print("A")
} else if (grade >= 80) {
  print("B")
} else if (grade >= 70) {
  print("C")
} else if (grade >= 60) {
  print("D")
} else {
  print("F")
}

```

Neat! We can also nest conditionals just like we do in Python.

```{r}


marks <- 55
attendance <- 80

if (marks > 50) {
  if (attendance > 75) {
    print("Passed")
  } else {
    print("Warning: Low Attendance")
  }
} else {
  print("Failed")
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}

numbers <- 1:5  

for (number in numbers) {
  print(paste("The current number is:", number))
}

counter <- 1

while (counter <= 5) {
  print(paste("Counter is at:", counter))
  counter <- counter + 1
}


```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}

v <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

print("Original vector:")
print(v)

indices_to_change <- c(2, 4, 5)

v[indices_to_change] <- v[indices_to_change] + 10

print("Modified vector:")
print(v)

```

We can loop over more than just lists:

```{r}
vector <- c(1, 2, 3, 4, 5)
for (item in vector) {
  print(item)
}

```

Like Python, we have while loops in R:

```{r}

counter <- 1

while (counter <= 5) {
  print(paste("Counter value:", counter))
  
  counter <- counter + 1
}

print("Loop finished.")

```

## Functions

When should we use functions?

```{r}

calculateArea <- function(radius) {
  area <- pi * radius^2
  return(area)
}

```

An example:

```{r}
area1 <- calculateArea(5)

area2 <- calculateArea(7)

print(area1)
print(area2)


```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}

result = main()
print(result)

```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}

df <- data.frame(
  A = rnorm(10),  
  B = rnorm(10),
  C = rnorm(10)
)

column_means <- apply(df, MARGIN = 2, FUN = mean)
print(column_means)

```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}

my_matrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2)

my_function <- function(x) {
    return(sum(x) + 10)
}

result_rows <- apply(my_matrix, 1, my_function)

result_cols <- apply(my_matrix, 2, my_function)

print("Result applying to rows:")
print(result_rows)
print("Result applying to columns:")
print(result_cols)

```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    String Reversal.

    ```{r}
    reverse_string <- function(s) {
      return(stringi::stri_reverse(s))
    }
    reverse_string("CodeWars") 
    #Simple Summation
    simple_sum <- function(n) {
      return(sum(1:n))
    }
    simple_sum(10) 
    #Find the Largest Number
    find_max <- function(numbers) {
      return(max(numbers))
    }
    find_max(c(1, 3, 5, 7, 9)) 
    #FizzBuzz
    fizzbuzz <- function(n) {
      for (i in 1:n) {
        if (i %% 15 == 0) {
          print("FizzBuzz")
        } else if (i %% 3 == 0) {
          print("Fizz")
        } else if (i %% 5 == 0) {
          print("Buzz")
        } else {
          print(i)
        }
      }
    }

    fizzbuzz(15) 

    rotate_array <- function(arr, n) {
      len <- length(arr)
      n <- n %% len  
      return(c(tail(arr, n), head(arr, len - n)))
    }
    rotate_array(c(1, 2, 3, 4, 5), 2) 
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    fizzBuzzCheck <- function(number) {
        if (number %% 3 == 0 & number %% 5 == 0) {
            return("FizzBuzz")
        } else if (number %% 3 == 0) {
            return("Fizz")
        } else if (number %% 5 == 0) {
            return("Buzz")
        } else {
            return(number)
        }
    }
    result <- fizzBuzzCheck(15)
    print(result) 

    result <- fizzBuzzCheck(9)
    print(result)  

    result <- fizzBuzzCheck(10)
    print(result)  
    result <- fizzBuzzCheck(7)
    print(result) 

    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    marioStairs <- function(n) {
      for (i in 1:n) {
        cat(strrep("X", i), "\n")
      }
    }
    marioStairs(4)

    ```
